Conversion from forest to agriculture in the Brazilian Amazon from 1985 to 2021

### 2022-12-05 manuscript/manuscript.Rmd
Many insertions to the text. Inclusion of conclusion. Improvement of results. Fix typos. Update figures.

abstract: |
  Land-use and land-cover changes in the Amazon biome are a main process that influences the environment and societies at local, national, and global levels. A large quantity of studies already relied on land-use classification to analyze this process. This study brings a new dataset created by calculating the time necessary for deforested areas to change to agriculture (annual and permanent crops) in the Brazilian Amazon biome. The calculations were performed over high resolution land-use classification by using simple algebra and recursion to identify every conversion from forest to agriculture, and ranges from 1985 to 2021. The new data can be useful in interdisciplinary studies about land-use and land-cover change analysis in Brazil, such as exploring its drivers and evaluating the effects on natural vegetation cover and ecosystem services. The major innovation brought by this study are links between the deforestation year of a given pixel with its respective year of agriculture establishment, which can provide new insights to understand long-term land-use conversion processes in tropical ecosystems.

@@

Land-use and land-cover (LULC) changes are known to cause impacts on different scales.
Deforestation and land use changes can affect hydrological processes in large basins [@Arias2018]; reduce the forest resilience to extreme events (e.g., drought) and other sources of perturbation [@Boulton2022]; deforestation and farming practices can reduce convective rainfall and increase surface temperature [@Maeda2021]; increase carbon emissions [@Gatti2021]; cause important impacts on fauna and flora diversity, soil properties and carbon stocks [@Nunes2022; @Rittl2017]; and even negatively impact public health [@Ellwanger2020].

Data on LULC classification are evolving rapidly in the last decade.
Initiatives such as MapBiomas [@Souza2020], launched in 2015, provides high resolution, annual classifications for the Brazilian territory.
Such types of data are essential to analyze LULC change processes, make causal inferences, and understand what drives and what are the impacts of the transformation of the landscape in Brazil.
In this context, this study aims to characterize and quantify the length of the conversions from forest formations to agriculture in the Amazon over the last 36 years.
The availability of this data can be useful to the development of interdisciplinary research involved with LULC changes.

@@

The first part of the processing was to create a table from the binary values data, and create unique identification for each pixel, their coordinates, calculate the area, and also address the municipality in which each pixel was inside.
In this part, a table was also generated with the spatial information of each tile, so that it is possible to convert the generated tables back to raster format.

The second part of the processing was to load each tile separately, convert it to a table and calculate every possible conversion from forest to agriculture (Figure \@ref(fig:routine-plot)).
The conversions were calculated pixel by pixel, where the LULC time series were extracted and analysed for each conversion cycle that was present for that pixel.
The calculation of the conversion length was extremely simple, it is the year of the establishment of agriculture subtracted by the last year classified as a forest, however, we add one year to the latter, so that it represents the year of deforestation (since it is more meaningful for analysis).

```{r routine-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Representation of the process to calculate the conversion length from forests to agriculture. All values presented are hiphothetical, and serves only as an ilustration of the method."}

@@

The process of calculation of the conversion length (Figure \@ref(fig:routine-plot)), can be breakdown to seven steps of processing:

@@

The steps 2 to 7 are performed recursively to identify multiple conversions, in case they are present.
In addition to the conversion length values and the years of the conversion, they are also qualified by the type of agriculture that were established after deforestation, if the conversion occurred from primary or secondary forests, and the number of occurences of conversions at the pixel.
The LULC classes within the conversion and until 5 years after were also stored.

@@

```{r tables-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = " Representation of the tables generated by the conversion length calculation. All values presented are hiphothetical, and serves only as an ilustration of the method."}

@@

The first type of table (Figure \@ref(fig:tables-plot).a) contain the spatial information (longitude, latitude and area) of each pixel, its unique identificators and the code of the municipality which contains the pixel.
The "cell_id" column contains a unique identification for the whole data, while the "tile_cell_id" holds a unique value of each pixel for a single tile.
Each tile is also uniquely identified in the column "tile_id".
This table is named as "mask_cells".

A second type of tables (Figure \@ref(fig:tables-plot).b) stores conversion length values, the first and last year of the conversion, the resulting agriculture type, and the number of the conversion cycle.
They also contain the unique id of each pixel and tile (related to the table above).
The "agri_code" column holds information about the type of agriculture the was established after deforestation.
The "forest_year" have values that represent the year of deforestation, the "agri_year" is the year when the agriculture establishment happened, and the subtraction between both (the conversion length) is stored in the "c_length" column.
Since it is possible that one pixels contains more than one conversion from forest to agriculture, we stored the number of the conversion in the column "c_cycle" (1 is the first conversion cycle).
However, conversion from forest to other LULC is also possible, so the column "f_cycle" holds information about the number of the conversion from forest to any LULC.
If the forest have suffered any conversion, it starts being considered as a secondary forest, as indicated in the column "forest_type".

@@

The tabular data was organized in a system of folders and sub folders in a hive format partitioning (Figure \@ref(fig:dataset-plot)).
This structure allows for easily access parts of the data set, and also filter it without the necessity to read the tables.
In the case of this data set, it is partitioned by the "tile_id" and "agri_cycle" variables.
The data set also have a metadata file with the spatial characteristics of each file, it stores the identification of the tile, the maximum and minimum latitude and longitude, the coordinate reference system, and the number of columns, rows, and cells of each tile.

```{r dataset-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = " Representation of the data set, structured in folders and sub folders."}

@@

To evaluate the accuracy, we calculated the Mean Absolute Error (MAE), the Bias (BIAS), and the Percent Bias (PBIAS).
We also analyzed the results by plotting the errors as frequency bars, and scatter plots between observed and estimated values.
After the completion of the analysis of the 100 sample points, we also conducted a qualitative assessment, where we compared our results with the satellite images.

@@

The conversion calculations show that between 1985 and 2021, 64,874 square kilometers of forests were converted to agriculture in the Brazilian Amazon biome.
The length of the conversions can go from 0 to 35 years, in which conversions closer to 0 years are considered as fast conversions, and conversions closer to 35 years are considered as slow conversions.
Transitions of 0 years are considered "direct" conversions, where there was no presence of pasture before the establishment of agriculture.
Our estimations show that around 9.2% of the conversions were "direct".

@@

```{r map-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Map of distribution of conversions from forest to agriculture in the Brazilian Amazon biome. The hexagonal cells represent the most common conversion length, and do not reflect the amount of area of conversions inside a cell. Transitions are concentrated in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The conversion length ranges from 0 (dark blue) to 35 years (light pink), and clusters of fast conversions (conversions closer to 1 year) can be discerned from clusters of slow conversions (conversions closer to 36 years). The histogram located in the bottom left shows that fast conversions are more common than slower conversions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different conversion lengths between properties. The red square shows the extent of the zoomed map."}

@@

Well defined clusters can be observed in the map created with aggregated conversion length data (Figure \@ref(fig:map-plot)).
Slow conversion areas tend to concentrate in specific regions of the biome, while fast conversions seem to have a wider distribution, but also tend to form spatial clusters.
However, this pattern does not hold completely when observing the data at its original scale (Figure \@ref(fig:map-plot)), where areas with different conversion lengths are mixed between each other.
When observing at the original scale, we could not spot any well defined pattern or direction of the occurrence of faster to slower conversions (Figure \@ref(fig:map-plot)).

Other studies investigating patterns of conversions in the Amazon also found the formation of clusters of patterns, although there is a big heterogeneity at larger scales [@MullerHansen2017].
The data can be analyzed year by year, and also be separated by primary and secondary forests being converted to agriculture (Figure \@ref(fig:transbar-plot)).

```{r transbar-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Transition area from forest to agriculture, per year and conversion length. The bars represent the total amount of area at some state of the conversion for each year. The color gradient in each bar represents the conversion length related to a deforestation or an agriculture establishment event. Blue tones represents fast conversions (closer to 0 years), pink tones represent slow conversions (closer to 35 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent agriculture establishment of primary forests (b) and secondary forests (d)."}

@@

The deforestation area of primary forests increased largely from 1986 to 2003, which was followed by a steep decrease until 2009, when the deforested areas reached a stable rate (Figure \@ref(fig:transbar-plot).a).
From 1986 to approximately 1995, most of the deforested areas suffered a slow conversion, mostly were higher than 10 years.
After this period, fast conversions started to become more common, specially from 2002 to 2004 (Figure \@ref(fig:transbar-plot).a).
The peak of deforestation in 2003, according to our results, shows that 12.6% of the conversions were a direct conversion from forest to agriculture.
Previous studies estimated a proportion of 23% [@Morton2006].
If we include fast conversions (from 0 to 2 years) from our results, the proportion jumps to 48.5 % of all conversions.

Agricultural establishment over areas of primary forests peaked in 2005 and 2013 (Figure \@ref(fig:transbar-plot).b).
Despite similar agricultural establishment rates between both years, their conversion lengths differ greatly, in 2005 most of the conversions were faster than 10 years, while in 2013 the great majority of conversions were slower than 10 years (Figure \@ref(fig:transbar-plot).b).
In 2003 it was observed a shift in the conversion length from forest until the establishment of agricultural areas.
From this year onward, most conversions happened in areas deforested at least 10 years before (Figure \@ref(fig:transbar-plot).b).
Even after the decrease of deforestation after 2002, areas under agriculture expanded over lands where deforestation happened before 2002 (mostly over pastureland).
However, after 2019, a sudden drop of agriculture establishment rate happened (Figure \@ref(fig:transbar-plot).b).

The period between 2001 and 2005 showed a peak of deforestation (Figure \@ref(fig:transbar-plot).a), and of croplands expansion (Figure \@ref(fig:transbar-plot).b).
From 2006 to 2010, we observed a decrease in both variables, a process that was already observed between 2001 and 2010 [@Macedo2012], which points out a decoupling of soybean expansion and deforestation.
However, our result shows that after 2011, agricultural areas rapidly expanded in the amazon biome, especially at areas deforested many years before.

@@

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soybean grains harvested from areas deforested after 2008.
Our estimates of conversions show a decrease of deforested areas after 2008, where it reached minimal values (Figure \@ref(fig:transbar-plot).a).
After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared more than a decade before (therefore, before 2006) (Figure \@ref(fig:transbar-plot).b) This shows that agriculture expansion did not halt after the soy moratorium, but producers expended over areas cleared many years ago, which may have indirect impacts on deforestation increase over distant areas [@Arima2011]. However, this indirect effect is not always clearly observed when analyzing the Amazon biome as a whole, but our data shows that periods of fast agriculture expansion occurred at the same time with minimal deforestation values, according to PRODES estimates (add reference). Expansion of agricultural areas also occurred over cleared areas of secondary forests in 2012, but with an important amount of fast conversions (Figure \@ref(fig:transbar-plot).d). The causes of the increase of agricultural establishment areas can be numerous, and one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests [@Kroger2017; @Pereira2019].

The conversion length patterns across years can change significantly between different states (Figure \@ref(fig:transridge-plot)).

```{r transridge-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Transition length patterns inside states in the Amazon biome. Each year has a density estimate of the conversion lengths, represented as colored curves. The peak of the curves represents conversion length values with more frequency in one year of one state. Blue tones represents fast conversions (conversions closer to 0 years), pink tones represents slow conversions (conversions closer to 35 years)"}

@@

```{r errorbar-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Bar with the count of error values (difference between observed and estimated values). Positive values indicate underestimation of the variable (estimations were lower than observations), negative values indicate overestimation (estimations higher than observations). Error metrics (Mean Absolute Error, Bias and Percent Bias) are displayed in the top right position of each box. The white dashed lines represent the MAE values of each variable."}

@@

```{r errorscatter-plot, out.width = "17cm", fig.pos = "ht", echo = FALSE, fig.cap = "Scatter plot between paired values of estimated and observed Deforestation and Agriculture Establishment years, and their respective Transition Length. The gray line represents values which there would be no errors of the estimated values in comparison with the observed values. Translucid white areas represent quantile ranges (5\\%, 25\\%, 75\\% and 95\\%) of the errors. The area between 5\\% and 95\\% quantiles includes 90\\% of points. The area between 25\\% and 75\\% quantiles includes 50\\% of points."}

@@

According to accuracy assessment from MapBiomas, the collection 7 presents a global accuracy of 96.6% for the Amazon biome, which is the proportion of pixels that were classified correctly.
For the Forest class, MapBiomas showed small errors of inclusion (proportion of pixels misclassified as other classes, but the real class were Forest), which fluctuated around 1%.
The omission errors for Forests are also small (proportion of pixels misclassified as Forests, but the real class were not Forest), which fluctuated around 2%.
The Agriculture class presented more errors, the inclusion errors ranged from 22% to 5%, and were mostly composed by Forests pixels (forest pixels misclassified as agriculture).
The omission errors of Agriculture ranged from 22% to 8%, and were also mostly composed of Forest pixels (agriculture pixels misclassified as forest).

@@

High heterogeneity inside agriculture plots (sample 56), caused by both year of deforestation or year of agriculture establishment.
Roads being considered as agriculture (sample 2).
Omissions were detected, specially agriculture plots not being considered as a conversion as a whole, parts of a plot that were clearly a conversion were not considered as a conversion (sample 16, 30, 75).
Inclusions were detected, areas that clearly showed no agriculture were considered as conversions (sample 10, 98).
Effect of border on results (sample 25, 56, 84, 92).
Areas with sparse forest vegetation presented a large error of deforestation year (sample 33).

Patterns of agricultural areas are roughly well represented (sample 25, 43, 56, 84).
Some locations present a better homogeneity of conversions inside agriculture plots (sample 84).

Interference due to atmospheric conditions and lack of image availability clearly shaped spatial patterns of conversions at some locations.
Quality of composites through the time series affected accuracy of results.
Regions of the North part of the Amazon present lower quality than regions in the South of the Amazon biome.

@@

>>>>

abstract: |
  Land-use and land-cover changes in the Amazon biome are main processes that influence the environment and societies at local, national, and global levels. A large quantity of studies already relied on land-use classification to analyze this process. This study brings a new dataset created by calculating the time necessary for deforested areas to change to agriculture (annual and permanent crops) in the Brazilian Amazon biome. The calculations were performed over high resolution land-use classification by using simple algebra and recursion to identify every conversion from forest to agriculture, and ranges from 1985 to 2021. The new data can be useful in interdisciplinary studies about land-use and land-cover change analysis in Brazil, such as exploring its drivers and evaluating the effects on natural vegetation cover and ecosystem services. The major innovation brought by this study are links between the deforestation year of a given pixel with its respective year of agriculture establishment, which can provide new insights to understand long-term land-use conversion processes in tropical ecosystems.

@@

Land-use and land-cover (LULC) changes are known to cause impacts on different scales.
Deforestation and LULC changes can affect hydrological processes in large basins [@Arias2018]; reduce the forest resilience to extreme events (e.g., drought) and other sources of perturbation [@Boulton2022]; deforestation and farming practices can reduce convective rainfall and increase surface temperature [@Maeda2021]; increase carbon emissions [@Gatti2021]; cause important impacts on fauna and flora diversity, soil properties and carbon stocks [@Nunes2022; @Rittl2017]; and even negatively impact public health [@Ellwanger2020].

Data on LULC classification are evolving rapidly in the last decade.
Initiatives such as MapBiomas [@Souza2020], launched in 2015, provides high resolution, annual classifications for the Brazilian territory.
Such types of data are essential to analyze LULC change processes, make causal inferences, and understand what drives and what are the impacts of the transformation of the landscape in Brazil.
In this context, this study aims to quantify and characterize the length of the conversions from forest formations to agriculture in the Amazon over the last 36 years.
The availability of this data can be useful to the development of interdisciplinary research, since it can provide useful insights into the environmental, economic, and social impacts of land conversion.
For example, this data can be used to develop insights into the economic costs and benefits associated with land conversion, the effects of land conversion on local ecosystems, and the impact of land conversion on livelihoods and social dynamics.

@@

The first part of the processing was to create a table from the binary values data, and create unique identification for each pixel, their coordinates, calculate the area, and also the municipality in which each pixel was inside.
In this part, a table was also generated with the spatial information of each tile, so that it is possible to convert the generated tables back to raster format.

The second part of the processing was to load each tile separately, convert it to a table and calculate every possible conversion from forest to agriculture (Figure \@ref(fig:routine-plot)).
The conversions were calculated pixel by pixel, where the LULC time series were extracted and analyzed for each conversion cycle that was present for that pixel.
The calculation of the conversion length was simple, it is the year of the establishment of agriculture subtracted by the last year classified as a forest, however, we add one year to the latter, so that it represents the year of deforestation (since it is more meaningful for analysis).

```{r routine-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Representation of the process to calculate the conversion length from forests to agriculture. All values presented are hyphothetical, and serve only as an illustration of the method."}

@@

The process of calculation of the conversion length (Figure \@ref(fig:routine-plot)), can be divided into seven steps of processing:

@@

The steps 2 to 7 are performed recursively to identify multiple conversions, in case they are present.
In addition to the conversion length values and the years of the conversion, they are also qualified by the type of agriculture that were established after deforestation, if the conversion occurred from primary or secondary forests, and the number of occurrences of conversions at the pixel.
The LULC classes within the conversion and until 5 years after were also stored.

@@

```{r tables-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = " Representation of the tables generated by the conversion length calculation. All values presented are hyphothetical, and serve only as an illustration of the method."}

@@

The first type of table (Figure \@ref(fig:tables-plot).a) contains the spatial information (longitude, latitude and area) of each pixel, its unique identification and the code of the municipality which contains the pixel.
The "cell_id" column contains a unique identification for the whole data, while the "tile_cell_id" holds a unique value of each pixel for a single tile.
Each tile is also uniquely identified in the column "tile_id".
This table is named "mask_cells".

A second type of tables (Figure \@ref(fig:tables-plot).b) stores conversion length values, the first and last year of the conversion, the resulting agriculture type, and the number of the conversion cycle.
They also contain the unique id of each pixel and tile (related to the table above).
The "agri_code" column holds information about the type of agriculture the was established after deforestation.
The "forest_year" have values that represent the year of deforestation, the "agri_year" is the year when the agriculture establishment happened, and the subtraction between both (the conversion length) is stored in the "c_length" column.
Since it is possible that one pixel contains more than one conversion from forest to agriculture, we stored the number of the conversion in the column "c_cycle" (1 is the first conversion cycle).
However, conversion from forest to other LULC is also possible, so the column "f_cycle" holds information about the number of the conversion from forest to any LULC.
If the forest has suffered any conversion, it starts being considered as a secondary forest, as indicated in the column "forest_type".

@@

The tabular data was organized in a system of folders and sub folders in a hive format partitioning (Figure \@ref(fig:dataset-plot)).
This structure allows for easily accessing parts of the data set, and also to filter it without the necessity to read the tables.
In the case of this data set, it is partitioned by the "tile_id" and "agri_cycle" variables.
The data set also has a metadata file with the spatial characteristics of each file, it stores the identification of the tile, the maximum and minimum latitude and longitude, the coordinate reference system, and the number of columns, rows, and cells of each tile.

```{r dataset-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = " Representation of the data set, structured in folders and sub folders."}

@@

To evaluate the accuracy, we calculated the Mean Absolute Error (MAE), the Bias (BIAS), and the Percent Bias (PBIAS).
We also analyzed the results by plotting the errors as frequency bars, and scatter plots between observed and estimated values.
After the completion of the analysis of the 100 sample points, we also conducted a qualitative assessment, where we compared our results with satellite images composites.

@@

The conversion calculations show that in the Brazilian Amazon biome, 64874 square kilometers of forests were converted to agriculture between 1985 and 2021.
The length of the conversions can go from 0 to 35 years, in which conversions closer to 0 years are considered as fast conversions, and conversions closer to 35 years are considered as slow conversions.
Transitions of 0 years are considered "direct" conversions, where there was no presence of pasture before the establishment of agriculture.
Our estimations show that around 9.2% of the conversions were "direct".

@@

```{r map-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Map of distribution of conversions from forest to agriculture in the Brazilian Amazon biome. The hexagonal cells represent the most common conversion length, and do not reflect the amount of area of conversions inside a cell. Transitions are concentrated in the south (Mato Grosso state), and in the east (Pará and Maranhão states). The conversion length ranges from 0 (dark blue) to 35 years (light pink), and clusters of fast conversions (conversions closer to 0 year) can be discerned from clusters of slow conversions (conversions closer to 35 years). The histogram located in the bottom left shows that fast conversions are more common than slower conversions. The zoomed map in the bottom right shows the results in finer resolution, where it is possible to observe different conversion lengths between properties. The red square shows the extent of the zoomed map."}

@@

Well defined clusters can be observed in the map created with aggregated conversion length data (Figure \@ref(fig:map-plot)).
Slow conversion areas tend to concentrate in specific regions of the biome, while fast conversions seem to have a wider distribution, but also tend to form spatial clusters.
However, this pattern does not hold completely when observing the data at its original scale (Figure \@ref(fig:map-plot)), where areas with different conversion lengths are mixed between each other.
When observing at the original scale, we could not spot any well defined pattern or direction of the occurrence of faster to slower conversions (Figure \@ref(fig:map-plot)).
Other studies investigating patterns of conversions in the Amazon also found the formation of clusters of patterns, although there is a big heterogeneity at larger scales [@MullerHansen2017].

The data can be analyzed year by year, and also be separated by primary and secondary forests being converted to agriculture (Figure \@ref(fig:transbar-plot)).

```{r transbar-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Transition area from forest to agriculture, per year and conversion length. The bars represent the total amount of area at some state of the conversion for each year. The color gradient in each bar represent the conversion length related to a deforestation or an agriculture establishment event. Blue tones represents fast conversions (closer to 0 years), pink tones represent slow conversions (closer to 35 years). Transition events were separated by deforestation of primary forests (a) and secondary forests (c), and the subsequent agriculture establishment of primary forests (b) and secondary forests (d)."}

@@

The deforestation area of primary forests increased largely from 1986 to 2003, which was followed by a steep decrease until 2009, when the deforested areas reached a stable rate (Figure \@ref(fig:transbar-plot).a).
From 1985 to 1995, most deforested areas suffered a slow conversion, mostly were higher than 10 years (76 %).
After this period, fast conversions started to become more common (Figure \@ref(fig:transbar-plot).a), specially from 2002 to 2004, where 70 % of the transitions were faster than 10 years.
Our results show that in 2003 (the peak of deforestation), 12.7% of the conversions were a direct conversion from forest to agriculture.
Previous studies estimated a proportion of 23 % in the Mato Grosso state, for the same year [@Morton2006].
If we include fast conversions (from 0 to 2 years) from our results, the proportion jumps to 48.1 % of all conversions in 2003.

Agricultural establishment over areas of primary forests peaked in 2005 and 2013 (Figure \@ref(fig:transbar-plot).b).
Despite similar agricultural establishment areas in both years, their conversion lengths differ greatly (Figure \@ref(fig:transbar-plot).b), in 2005 most of the conversions were faster than 10 years (78 %), while in 2013 the majority of conversions were slower than 10 years (66 %).
In 2010, it was observed a shift in the conversion length from forest until the establishment of agricultural areas.
From this year onward, 68 % of the conversions happened in areas deforested at least 10 years before (Figure \@ref(fig:transbar-plot).b).
Even after the decrease of deforestation after 2002, areas under agriculture expanded over lands where deforestation happened before 2002 (mostly over pastureland).
However, after 2019, a sudden drop of agriculture establishment rate happened (Figure \@ref(fig:transbar-plot).b).
This behavior could be related to economic causes or even the scarcity of suitable areas for agricultural expansion, which seems unlikely, since the last years have been marked by an increase of deforestation rates and the advance of pastureland over the forest.

@@

In 2006, the Brazilian Association of Vegetable Oil Industries (BIOVE) and the National Association of Cereal Exporters (ANEC) committed to avoid commercialization of soybean grains harvested from areas deforested after 2008, known as the soy moratorium.
The soy moratorium is considered to have contributed to the decline of the deforestation directly related to soybean expansion [@Paim2021; @Amaral2021; @Heilmayr2020; @Kastens2017]. Our estimates of conversions show a decrease of deforested areas after 2008, where it reached minimal values (Figure \@ref(fig:transbar-plot).a). After 2006, agriculture establishment over deforested primary forests suffered a decrease, which stayed relatively stable until 2012, where a steep increase occurred, however, the new areas being occupied by agriculture were mainly over areas that were cleared before 2008 (Figure \@ref(fig:transbar-plot).b). The period between 2006 and 2010 is also considered as a turnover point, known as the decoupling of deforestation and soybean production, where the total production was not as dependent on agriculture expansion as before [@Macedo2012]. Therefore, the observed increase of agricultural establishment after 2012 shows that agriculture expansion did not halt after the soy moratorium nor after periods of high soybean productivity. Producers started to expand over areas cleared many years ago, which may still have indirect impacts on deforestation increase over distant areas [@Arima2011; @Gollnow2018]. However, this indirect effect is not always clearly observed when analyzing the Amazon biome as a whole, our data shows that periods of fast agriculture expansion occurred at the same time with minimal deforestation values, according to PRODES estimates [@Assis2019]. Expansion of agricultural areas also occurred over cleared areas of secondary forests in 2012, but with an important amount of fast conversions (Figure \@ref(fig:transbar-plot).d). The causes of the increase of agricultural establishment areas can be numerous, and one main driver was the approval of a new Forest Code, in 2012, which is considered to have undermined the environmental protection of forests [@Kroger2017; @Pereira2019], since it opened the possibility of amnesty for illegally deforested areas [@Schielein2018; @Santanna2021; @Filho2014], and the possible reduction of legal reserve area [@Freitas2018].

Even though it is possible to analyze the results in the Amazon as a whole, the conversion length patterns across years can change significantly between different states (Figure \@ref(fig:transridge-plot)).

```{r transridge-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Transition length patterns inside states in the Amazon biome. Each year has a density estimate of the conversion lengths, represented as colored curves. The peak of the curves represents conversion length values with more frequency in one year of one state. Blue tones represent fast conversions (conversions closer to 0 years), pink tones represent slow conversions (conversions closer to 35 years)."}

@@

```{r errorbar-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Bar with the count of error values (difference between observed and estimated values). Positive values indicate underestimation of the variable (estimations were lower than observations), negative values indicate overestimation (estimations higher than observations). Error metrics (Mean Absolute Error, Bias and Percent Bias) are displayed in the top right position of each box. The white dashed lines represent the MAE values of each variable."}

@@

```{r errorscatter-plot, out.width = "17cm", fig.pos = "h", echo = FALSE, fig.cap = "Scatter plot between paired values of estimated and observed Deforestation and Agriculture Establishment years, and their respective Transition Length. The gray line represents values where there would be no errors of the estimated values in comparison with the observed values. Translucid white areas represent quantile ranges (5\\%, 25\\%, 75\\% and 95\\%) of the errors. The area between 5\\% and 95\\% quantiles includes 90\\% of points. The area between 25\\% and 75\\% quantiles includes 50\\% of points."}

@@

According to accuracy assessment from MapBiomas, the collection 7 presents a global accuracy of 96.6% for the Amazon biome, which is the proportion of pixels that were classified correctly.
For the Forest class, MapBiomas showed small errors of inclusion (proportion of pixels misclassified as other classes, but the real class were Forest), which fluctuated around 1%.
The omission errors for Forests are also small (proportion of pixels misclassified as Forests, but the real class were not Forest), which fluctuated around 2%.
The Agriculture class presented more errors, the inclusion errors ranged from 22% to 5%, and were mostly composed of Forests pixels (forest pixels misclassified as agriculture).
The omission errors of Agriculture ranged from 22% to 8%, and were also mostly composed of Forest pixels (agriculture pixels misclassified as forest).

@@

The qualitative assessment of 100 samples was performed to analyze the results in conjunction with the error calculations and metrics.
It allows the interpretation of results with more context, since results are compared visually with reference satellite image composites, in an area of 4 square kilometers for each sample.

We found high heterogeneity inside agriculture plots (sample 56), caused by both year of deforestation or year of agriculture establishment.
This heterogeneity may be caused by smaller uncertainties within properties, where it would be expected higher homogeneity of conversion length values.
In few samples, roads were considered as agriculture (sample 2), so there was in fact no conversion to agriculture.
Other inclusion errors were detected, where areas that clearly showed no agriculture were considered as conversions (sample 10, 98).
There were also omissions errors, especially agriculture plots not being considered as a conversion as a whole, where parts of a plot that were clearly a conversion to agriculture were not considered as one (sample 16, 30, 75).
A lot of uncertainty was identified at borders of agricultural areas (sample 25, 56, 84, 92), which is a very common issue in classification of satellite imagery, since there are the presence of mixed pixels, where different objects share the space of a same pixel [@Kaur2019].
We also observed that areas with sparse forest vegetation presented a large error of deforestation year (sample 33).

We also observed that in many samples, the patterns of agricultural areas were roughly well represented (sample 25, 43, 56, 84), the shape of agricultural plots were clear and it was possible to distinguish different agricultural areas.
Some locations present a higher homogeneity of conversions inside agriculture plots (sample 84), which may be related to areas with smaller uncertainties.

@@

The conversion from forests to agriculture estimates can be a valuable source of information to support research over different disciplines related to LULC changes.
We showed that our data relates to many historical events in Brazil, which shaped the landscape of the Brazilian Amazon biome, covering public policies, economic and environmental aspects.
The data is structured and stored in a ready to analyze format, which is very convenient to be analyzed in conjunction with economic and demographic data.

Our data also explores new aspects of the MapBiomas classification data, which was not yet analyzed, by explicitly linking deforestation with agricultural establishment.
The accuracy assessment creates new opportunities for future improvements of LULC data, especially concerning conversions from forest to agriculture.
